---
title: "An analysis of casualties in fire accidents in Bangladesh 2020-2022"
author: "Md. Sourav Hasan Rabby"
date: "May 23, 2023"
output:

  html_document:
    toc: yes
    toc_depth: 4
    fig_width: 10
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is a simple analysis of casualties occurred in a number of fire accidents in Bangladesh in between year 2020 and 2022. The data has been taken from Bangladesh Fire Service [website](fireservice.gov.bd). This notebook also explains how the analysis was done.

## Why choosing 2020-2022 timeline?

Though in the [website](fireservice.gov.bd) of ***Bangladesh Fire Service*** there have been data from 2015 to 2022, the format of the documents from 2015-2019 is structurally different from those of 2020-2022. As a result they can't be merged or can be merged after lots of modification. We have a plan to make a separate analysis for 2015-2019 timeline.

## Desclaimer

This analysis is a personal project and no in any way suggests or claims anything.

## Analysis

#### Importing libraries

We at first import necessary libraries for analysis.

```{r library, echo=TRUE}

library(tidyverse) # for data cleaning, wrangling and vizualization.
library(treemapify) # for treemap graphical vizualization.


```

#### Importing raw data

Now let import raw dataset. You can get the raw data [here](https://drive.google.com/drive/folders/1QzjEYWuTmctMSHcBCbo1vGdkep0gb7qV). We also look at the head of the dataset and get the column names of the dataset.

```{r data import, echo=TRUE}
data <- read.csv("E:/Fire Data analysis/Data/Fire injury death 2020-2022.csv")
head(data)
colnames(data)
```

#### Data cleaning and modification

We don't need the columns which deals with 'total' of other columns (for example: total_common_injured column can be obtained from the summation of two prior columns.). Also we won't deal with the 'rescue' columns. So we will scrap those columns.

There is a row named **total** which shows the summation of each column which we don't need since we can compute those easily. So we will delete this row too.

```{r}
# Removing specific columns having total and rescued data
data2 <- data |> select(1, 2, 3, 4, 6, 7, 9, 10)
data2 <- data2[-40,] # Removing 'total' row
```

Let us look at the dataset now. We will look at the column names too.

```{r}
head(data2)
colnames(data2)
```

We will now rename the columns which begins with *'common_poeple\_'* to give those columns with shorter names. We also will rename the column with term *'fire_employee'* to **fire employee**.

```{r}
colnames(data2) <- gsub('common_people_', '', colnames(data2))
colnames(data2) <- gsub('fire_employee', 'fire employee', colnames(data2))
```

Let us check the column names now:

```{r}
colnames(data2)
```

Now for the sake of analyzing easily, we will move the rows with nature of casualty (injured or dead) and type of the victims (male, female of fire employee) to two new columns with name **State** and **Type** respectively.

```{r}
fire_data_x <- data2 |>
                pivot_longer(
                    cols = c(3, 4, 5, 6, 7, 8),
                    names_to = c("State", "Type"),
                    names_sep = "_",
                    values_to = "Casualty"  

                )
head(fire_data_x)
```

Now we will move the **State** column to two different columns named **injured** and **dead** so that each row of the dataset shows the number of injured or dead victims for each fire accident scene and victim type.

```{r}
fire_data <- fire_data_x |> 
  pivot_wider(
    names_from = State,
    values_from = Casualty
  )
head(fire_data)
```

We will use [**fire_data**]{style="color:blue"} for most of our future analyses.

#### Yearly casualty

Now let us check what are the numbers of casualty per year.

```{r}
yearly_casualties <- fire_data |>
                    group_by(year) |>
                        summarize(
                            death = sum(dead),
                            injury = sum(injured)
                        )
yearly_casualties 
```

```{r include=FALSE}
total_casualty <- fire_data |> 
                      summarize(
                        Type = "Total casualty",
                        Injured = sum(injured),
                        Dead = sum(dead)
                      )
```

```{r echo=TRUE}
yearly_casualty_plot <- yearly_casualties |> 
                    pivot_longer(
                    cols = c(2, 3),
                    names_to = c("State"),
                    values_to = "Count") |>
                    ggplot(aes(x = year, y = Count, fill = State))+
                    geom_bar(stat = "identity", position = "dodge")+
                    labs(
                          x = "Year",
                          y = "Nos. of casualty",
                          fill = "Casualty type",
                          title = "Fire accident casualty 2020-2022",
                          subtitle = "By year",
                          caption = "Data from fireservice.gov.bd"
                      )+
                    theme(
                          panel.background = element_rect(fill = "light gray"),
                          plot.title = element_text(size = 20),
                          plot.subtitle = element_text( size = 15),
                          plot.caption = element_text(size = 10, face = "italic"),
                          axis.title.x = element_text(size = 13, vjust = -5),
                          axis.title.y = element_text(size = 13, vjust = 2),
                          axis.text = element_text(size = 15)
                      )
```

```{r yearly_casualty_plot, echo= FALSE, fig.align = "center", fig.cap = "fig 1: Fire accident casualty 2020-2022: By year"}
plot(yearly_casualty_plot)
```

It seems that 2021 was the most devastating year in terms of casualties. 219 people were dead and 576 people were injured in fire accidents in that year. It is also seen that a total number of people `r total_casualty[1,2]` died and `r total_casualty[1,2]` injured between 2020-2022 due to fire accidents.

#### Casualty by victim type

Now let us check casualty by victim type. As mentioned earlier, the dataset has divided victims into 3 type: *male*, *female* and *fire employee*. Please be noted that here the types male and female doesn't include male and female fire department employees.

```{r}
casualties_by_type <- fire_data |>
                            group_by(Type) |>
                                summarize(
                                    Injured = sum(injured),
                                    Dead = sum(dead)
                                ) |>
                              mutate(Total = rowSums(across(where(is.numeric))))
casualties_by_type
```

```{r}
casualty_by_type_plot <- casualties_by_type |>
            pivot_longer(
                    cols = c(2, 3),
                    names_to = c("State"),
                    values_to = "Count"   

            ) |>
            ggplot(aes(Type, Count, fill = State))+
            geom_bar(stat = "identity", position = "dodge")+
            labs(
                  x = "Victim type",
                  y = "Nos. of casualty",
                  fill = "Casuality type",
                  title = "Fire accident casualty 2020-2022",
                  subtitle = "By gender or employment",
                  caption = "Data from fireservice.gov.bd"
            )+
            theme(
                  panel.background = element_rect(fill = "light blue"),
                  plot.title = element_text(size = 20),
                  plot.subtitle = element_text( size = 15),
                  plot.caption = element_text(size = 10, face = "italic"),
                  axis.title.x = element_text(size = 13, vjust = -5),
                  axis.title.y = element_text(size = 13, vjust = 2),
                  axis.text = element_text(size = 15)
            )
```

```{r casualty_by_type_plot, echo= FALSE, fig.align = "center", fig.cap = "fig 2: Fire accident casualty 2020-2022: By victim type"}
plot(casualty_by_type_plot)
```

It is seen that though the number of female victim injured is less than one third that of male victim, surprisingly the fatality count of female is almost same to male. Let us check number of male, female and fire employee victims by year.

#### Casualty by type per year

```{r include=FALSE}
casualty_by_type_by_year <- fire_data |>
  group_by(Type, year) |>
  summarize(
    Injured = sum(injured),
    Dead = sum(dead)
  )
casualty_by_type_by_year
```

```{r}
casualty_by_type_by_year_plot <- casualty_by_type_by_year |>
                  ggplot(aes(x = year, y = Dead, fill = Type ))+
                  geom_col(position = "dodge")+
                  labs(
                  x = "Year",
                  y = "Nos. of casualty",
                  fill = "Victim type",
                  title = "Fire accident casualty 2020-2022",
                  subtitle = "By victim type per year",
                  caption = "Data from fireservice.gov.bd"
            )+
            theme(
                  panel.background = element_rect(fill = "light blue"),
                  plot.title = element_text(size = 20),
                  plot.subtitle = element_text( size = 15),
                  plot.caption = element_text(size = 10, face = "italic"),
                  axis.title.x = element_text(size = 13, vjust = -5),
                  axis.title.y = element_text(size = 13, vjust = 2),
                  axis.text = element_text(size = 15)
            )
```

```{r casualty_by_type_by_year_plot, echo= FALSE, fig.align = "center", fig.cap = "fig 3: Fire accident casualty 2020-2022: By victime type per year"}
plot(casualty_by_type_by_year_plot)
```

It is seen in the above bar chart that in 2020, the death count of female due to fire accidents were very much higher than that of male. In 2021, both the bars for female and male are high. 2022 shows a reverse trend than that of 2020, male death count is much higher than female death count. It's an interesting observation. We have to check what happened in 2020 and 2022 that showing reverse trend. We also have to check why the casualty rate(both injury and death) at 2021 is so higher for both male and female compare to other years. It is also noticeable that 2022 is the only year with fire employee fatality.

Now let us check the casualty account for different fire scenes:

#### Casualty by fire accident scene

```{r}
casualty_scene <- fire_data |>
                      group_by(Description.of.fire.accident.scene)|>
                      summarize(
                        Injured = sum(injured),
                        Dead = sum(dead)
                      )
casualty_scene
```

Now let us check for the top five fire scenes where most of the death and injury occurred. We will left out *other scene fire* property type since it is a summation of other smaller types.

```{r}
# Top five fire scene with most number of death 
top_five_death <- casualty_scene |> 
  filter(Description.of.fire.accident.scene != "other scene fire") |>
    arrange(desc(Dead)) |> 
      slice(1:5) |>
        select(!Injured)
top_five_death
```

```{r}
top_five_death_plot <- top_five_death |>
  ggplot(aes(area = Dead, fill = Description.of.fire.accident.scene, 
             label = paste(Description.of.fire.accident.scene, Dead, sep = "\n")))+
  geom_treemap()+
  geom_treemap_text(
    colour = "white",
    place = "centre",
    size = 15        
  )+
  labs(
    title = "Fire accident death 2020-2022",
    subtitle = "By accident scene (top five)",
    caption = "Data from fireservice.gov.bd"
  )+
  theme(
    legend.position = "none",
    plot.title = element_text(size = 20),
    plot.subtitle = element_text( size = 15),
    plot.caption = element_text(size = 10, face = "italic")
  )
```

```{r top_five_death_plot, echo= FALSE, fig.align = "center", fig.cap = "fig 4: Fire accident casualty 2020-2022: Top five fire scene with most death"}
plot(top_five_death_plot)
```

```{r}
# Top five fire scene with most number of injury (except "other scene fire")
top_five_injury <- casualty_scene |> 
  arrange(desc(Injured)) |>
    slice(1:5) |>
      select(!Dead)
top_five_injury
```

```{r}
# Bar chart for top five injury scene
top_five_injury_plot <- top_five_injury |>
  ggplot(aes(reorder(Description.of.fire.accident.scene, Injured), Injured))+
  geom_col()+
  coord_flip()+
  labs(
                  x = "Fire accident scene",
                  y = "Injury count",
                  title = "Fire accident injury 2020-2022",
                  subtitle = "By accident scene (top five)",
                  caption = "Data from fireservice.gov.bd"
            )+
            theme(
                  panel.background = element_rect(fill = "light blue"),
                  plot.title = element_text(size = 20),
                  plot.subtitle = element_text( size = 15),
                  plot.caption = element_text(size = 10, face = "italic"),
                  axis.title.x = element_text(size = 13, vjust = -1),
                  axis.title.y = element_text(size = 13, vjust = 2),
                  axis.text = element_text(size = 15)
            )
```

```{r top_five_injury_plot, echo= FALSE, fig.align = "center", fig.cap = "fig 5: Fire accident casualty 2020-2022: Top five fire scene with most injury"}
plot(top_five_injury_plot)
```

So, it is seen that *factory-industry fire*, *other building fire* and *market-hat-bazaar-shop fire* are the places with most casualty since these three are common in both death and injury table. It is also noticeable that a lot of people die in vehicle fire and gas related fires (*gas cylinder fire* and *gas distribution pipeline fire*) causes a lots of injury.

#### Fatality by victim type and fire accident scene

Now let us check the fatality rate by victim type due to fire accidents in 2020-2022 timeline.

```{r include=FALSE}
total_fatality <- fire_data |>
                filter(Description.of.fire.accident.scene != "other scene fire") |>
                group_by(Description.of.fire.accident.scene, Type) |>
                summarize(
                          Total_fatality = sum(dead),
                          ) |>
                arrange(desc(Total_fatality))
                                
                                

total_fatality
```

From the above table we can see that both for male and female *factory-industry fire* and *vehicle fire* are the most fatal ones. Let us check which fire scenes are the most fatal for all type of victims

```{r}
max_fatality <- total_fatality |> 
                        ungroup() |>
                        filter(
                        Total_fatality == max(Total_fatality), .by = Type
                        )
max_fatality
```

It is clearly evident that the fire scenes most fatal both for male and fire employee are *factory-industry fire*s. On the other hand *vehicle fire*s are the most fatal ones for females.

#### Comparison of number of fire accidents with number of casualties

Now, let us compare the trend of the number of casualties with that of number of fire accidents during 2020-2022. For the number of fire accidents we will use a customized dataset made from datasets about property loss and rescue given in the [website](fireservice.gov.bd) of ***Bangladesh Fire Service***. The dataset can be found [here](https://drive.google.com/file/d/1KlJeTw-uleAMi-hQzJ7bRppPNjAx6AcO/view?usp=sharing)

*N.B* : We will try to provide another data analysis regarding that dataset which will discuss about monetary loss for these fire accidents.

```{r}
fire_count_x<- read.csv("E:/Fire Data analysis/Data/Fire accident count 2020-2022.csv")
head(fire_count_x)
colnames(fire_count_x)
```

Let us remove the column **description_of_cause** since it is not needed in our analysis. We will also remove the **Total** row.

```{r}
fire_count <- fire_count_x |>
                    filter(year != "Total") |>
                    select(year, count_of_fire_hazards) |>
                    group_by(year) |>
                    summarize(Total_count_of_accidents = sum(count_of_fire_hazards))
fire_count
```

Now let us join *yearly_casualties* tibble with *fire_count*. We will use <u>Left join</u> method here.

```{r}
yearly_data <- left_join(
                    fire_count,
                    yearly_casualties,
                    by = "year",
                   
                       ) 
yearly_data
```

This table shows us that the number of accidents has an upward trend while casualty count has a mixed trend. While 2021 was more fatal than 2020, there is significantly less number of casualty in 2022. Let's see the bar charts for easier comparison:

```{r include=FALSE}
yearly_data2 <- yearly_data |>
                group_by(year) |>
                summarize(
                  accident = Total_count_of_accidents / 10,
                  death = death,
                  injury = injury
                )
yearly_data2
```

```{r include=FALSE}
comparison_plot <- yearly_data2 |>
                    pivot_longer(
                      cols = c(2, 3, 4),
                      names_to = "variable",
                      values_to = "count"
                    ) |>
                    ggplot(aes(year, count,  fill = variable))+
                    geom_col(position = "dodge")+
                    labs(
                  x = "Year",
                  y = "Incident count",
                  fill = "",
                  title = "Fire accident casualty 2020-2022",
                  subtitle = "Accident and casualty count comparison",
                  caption = "Data from fireservice.gov.bd"
            )+
            theme(
                  panel.background = element_rect(fill = "light blue"),
                  plot.title = element_text(size = 20),
                  plot.subtitle = element_text( size = 15),
                  plot.caption = element_text(size = 10, face = "italic"),
                  axis.title.x = element_text(size = 13, vjust = -5),
                  axis.title.y = element_text(size = 13, vjust = 2),
                  axis.text = element_text(size = 15),
                 
                  
            )
```
```{r comparison plot, fig.align = 'center', fig.cap = "fig 6: Accident and casualty count comparison.                                                                                    Accident counts has been shown by a descrease of factor of 10 for better visualization" }
plot(comparison_plot)
```
From the bar chart we see that though there is a slight increase in the number of fire accidents in 2021, the number of casualty increased disproportionately. But on the other hand even though the number of fire accidents has increased in 2022 than those of both 2020 and 2021, its number of death is lower than the fatality in both 2020 and 2021; injury count is less than that of 2021 and slightly higher than the injury count of 2020. This is an interesting observation in our opinion.  